---
title: "ngme: an R package for non-Gaussian longitudinal data analysis"
author: "Asar, Bolin, Diggle, Wallin (alphabetical)"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Modelling framework 

The modelling frameworks that the package `ngme` provides functions for
has the following forms:
$$
\begin{aligned}
Y_{ij} = {\bf x}_{ij}^{\prime} {\bf \alpha} + {\bf d}_{ij}^{\prime} {\bf U}_i + Z_{ij},
\end{aligned}
$$
and
$$
\begin{aligned}
Y_{ij} = {\bf x}_{ij}^{\prime} {\bf \alpha} + {\bf d}_{ij}^{\prime} {\bf U}_i + W_i(t_{ij}) + Z_{ij}.
\end{aligned}
$$
Here, 

- $Y_{ij}$ is the $j$th response for subject $i$ at time $t_{ij}$,
- ${\bf x}_{ij}$ is the matrix of fixed effects explanatory variables,
- ${\bf \alpha}$ is the matrix of fixed effects coefficients,
- ${\bf d}_{ij}$ is the matrix of random effects explanatory variables,
- ${\bf U}_i$ is the matrix of subject-specific and time-invariant random effects coefficients,
- $W_i(t)$ is subject- and time-spceific random effect, specificed as a stochastic process,
- $Z_{ij}$ is random error.

`ngme` includes `nglda_est` and `nglda_predict` as the main functions 
 for parameter estimation based on maximum likelihood 
 implemented with a computationally efficient stochastic gradient algorithm and 
 prediction based on Gibbs sampling, respectively. `print`, `summary`, `plot`, 
 `fitted`, `residuals`, `intervals` functions are available as the supplemetary. 
 
 Within `ngme` one can choose to specify the distributions as
 
 - Normal and Normal-inverse Gaussian for ${\bf U}_i$,
 - Normal, Normal-inverse Gaussian, generalised asymmetric Laplace and Cauchy for $W_i(t)$,
 - Normal, Normal-inverse Gaussian and t for $Z_{ij}$.

# Data

```{r, echo=TRUE}
library(ngme)
data(srft_data)
knitr::kable(head(srft_data))
```

# Analysis

```{r, echo=TRUE}
set.seed(123)
rs_id <- sample(unique(srft_data$id), 100, replace = F)
srft_data_sub <- srft_data[srft_data$id %in% rs_id, ]

srft_fit <- nglda_est(
  fixed = log(egfr) ~ bage + sex + fu + pwl,
  random = ~ fu + bage|id,
  data = srft_data_sub,
  reffects = "NIG",
  process = c("Normal", "matern"),
  error = "tdist",
  timeVar = "fu",
  use.process = T,
  estimation.controls = list(nIter = 10, nIter.fisher = 10),
  other.controls = list(estimate.fisher = T, silent = T)
)

 #srft_fit2 <- update(srft_fit, use.process = F)

 srft_fit
 summary(srft_fit)
 head(fitted(srft_fit), 10)
 head(residuals(srft_fit), 10)
 intervals(srft_fit)

res.pre <- nglda_predict(srft_fit, id = unique(srft_data_sub$id), type = "Filter")
res.pre
plot(res.pre, id = unique(srft_data_sub$id)[11], pch = 19, cex = 0.5)
plot(res.pre, id = 2203400, pch = 19, cex = 0.5)
```

Vignettes are long form documentation commonly included in packages. Because they are part of the distribution of the package, they need to be as compact as possible. The `html_vignette` output type provides a custom style sheet (and tweaks some options) to ensure that the resulting html is as small as possible. The `html_vignette` format:

- Never uses retina figures
- Has a smaller default figure size
- Uses a custom CSS stylesheet instead of the default Twitter Bootstrap style

## Vignette Info

Note the various macros within the `vignette` section of the metadata block above. These are required in order to instruct R how to build the vignette. Note that you should change the `title` field and the `\VignetteIndexEntry` to match the title of your vignette.

## Styles

The `html_vignette` template includes a basic CSS theme. To override this theme you can specify your own CSS in the document metadata as follows:

    output: 
      rmarkdown::html_vignette:
        css: mystyles.css

## Figures

The figure sizes have been customised so that you can easily put two images side-by-side. 

```{r, fig.show='hold'}
plot(1:10)
plot(10:1)
```

You can enable figure captions by `fig_caption: yes` in YAML:

    output:
      rmarkdown::html_vignette:
        fig_caption: yes

Then you can use the chunk option `fig.cap = "Your figure caption."` in **knitr**.

## More Examples

You can write math expressions, e.g. $Y = X\beta + \epsilon$, footnotes^[A footnote here.], and tables, e.g. using `knitr::kable()`.

```{r, echo=FALSE, results='asis'}
knitr::kable(head(mtcars, 10))
```

Also a quote using `>`:

> "He who gives up [code] safety for [code] speed deserves neither."
([via](https://twitter.com/hadleywickham/status/504368538874703872))
